Understanding the nanoscale structure of hexagonal phase lyotropic liquid crystal membranes

### 2018-02-27 Ben_Manuscripts/structure_paper/Draft.tex
Some more comments.

\title{Understanding the nanoscale structure of hexagonal phase lyotropic
liquid crystal membranes}
\author{Benjamin J. Coscia}
\author{Michael R. Shirts}
\email{michael.shirts@colorado.edu}
\author{Richard D. Noble}
\author{Douglas L. Gin}
\affiliation{Department of Chemical and Biological Engineering, University of Colorado Boulder, Boulder, CO 80309, USA}
\alsoaffiliation{Department of Chemistry and Biochemistry, University of Colorado Boulder, Boulder CO 80309, USA}
\author{Joe Yelk}
\author{Matthew Glaser}
\affiliation{Department of Physics, University of Colorado Boulder, Boulder CO, 80309, USA}
\author{Xunda Feng}
\affiliation{Department of Chemical and Environmental Engineering, Yale University, New Haven, Connecticut 06511, USA}

\begin{document}

@@

@@

  Crosslinked lyotropic liquid crystal (LLC) membranes may be capable of
  performing highly selective separations. LLCs are amphiphilic molecules that
  have the ability to self-assemble into porous nanostructures
  \cite{smith_ordered_1997} which can be crosslinked to create mechanically
  strong membrane films with pores on the order of 1 nm in diameter
  \cite{zhou_supported_2005}. Unlike commercial NF membranes, LLC membrane pores
  are uniform in size. Since LLC membranes lack a pore size distribution, they
  inherently exhibit high selectivity due to their strict molecular weight
  cut-off (MWCO) ~\cite{zhou_supported_2005}. Additionally, the LLC monomers
  chosen to make membranes for this application are salts, and therefore lead to
  Donnan exclusion of ions in solution. The membrane gains a net surface charge
  when counterions from head groups that line the pore walls escape into the feed
  solution in order to balance the gradients of concentration and electric
  potential \cite{donnan_theory_1995}.

@@

  The H\textsubscript{II} phase pore geometry is better for transport than the
  Q\textsubscript{I} phase. The H\textsubscript{II} phase forms at room
  temperature in the presence of c.a.~10 wt\% water and consists of hexagonally
  packed, hydrophilic pore columns\cite{smith_ordered_1997}. In the absence of
  water, neat monomer will form the same hexagonal columnar structure which, in
  literature, has been referred to as the Col\textsubscript{h} thermotropic
  phase\cite{feng_scalable_2014} (See Figure~\ref{fig:hII_phase}).  The most
  promising Q\textsubscript{I} phase for membrane applications forms at 70\degree
  C when monomer is mixed with c.a.~20 wt\%
  glycerol\cite{carter_glycerol-based_2012}. Q\textsubscript{I} phase membranes
  consist of a tortuous network of three dimensionally interconnected pores that
  prevent optimal through-plane transport. The densely packed, non-tortuous and
  uniform sized pores of H\textsubscript{II} phase membranes represents the ideal
  geometry for achieving high solute flux\cite{matyka_tortuosity-porosity_2008}.
  Despite the promise of the H\textsubscript{II} phase, the hexagonally packed
  liquid crystalline domains, formed during self-assembly of NA-GA3C11 monomers, are
  isotropically aligned which is detrimental to membrane performance.

@@

  Our current understanding of LLC membrane systems is not rich enough to be
  able to precisely design them for specific separations. Over the past 20 years,
  H\textsubscript{II}-phase LLC polymer membrane studies have been limited
  primarily to Na-GA3C11 with some characterization done after minor structural
  modifications. Resel et al.~varied the length of the monomer tails and the
  counterion used and observed its affect on pore spacing
  \cite{resel_structural_2000}. In a later study of rejection performance, it was
  shown that membranes formed by cross-linked Na-GA3C11 in the
  H\textsubscript{II} phase cannot separate solutes less than 1.2 nm in diameter
  because the pores are too large \cite{zhou_supported_2005}. We do not yet
  understand how to controllably reduce the effective pore size or how to tune
  the chemical environment in the nanopores of this or related materials for
  small molecule separations. The only source of predictive modeling for LLC
  systems have been macroscopic models that likely do not adequately describe
  transport at these length scales \cite{hatakeyama_water_2011}. It will be
  challenging to efficiently narrow down the large design space in a laboratory
  setting without a robust model.

@@

  \begin{figure}
  \centering
	\begin{subfigure}{0.45\linewidth}
		\centering
		\includegraphics[width=\textwidth]{cartoon_pore.png}
		\caption{}~\label{fig:undetailed_pore}
	\end{subfigure}
	\begin{subfigure}{0.45\linewidth}
		\centering
		\includegraphics[width=\textwidth]{detailed_pore.png}
		\caption{}~\label{fig:detailed_pore}
	\end{subfigure}
  \caption{We can only speculate about solute behavior inside the membrane
	  pores based on our previous picture of the pore structure (a). We will use a
	  detailed molecular model in order to appropriately model the pore's complex
	  architecture which is crucial to understanding the mechanism of solute
	  transport (b). The head group region is colored blue
	  and the tail region is colored cyan in both representations.}~\label{fig:detail}
  \end{figure}

@@

  \item If layers do exist, how many monomers constitute a single layer? \label{point:monomernum}

  A simple molecular simulation study of a similar molecule suggested that
  there are 4 monomers in each layer \cite{zhu_methacrylated_2006}. Their model
  likely does not sufficiently imitate the chemical environment present in the
  real system since it only contains 16 total monomers. A separate calculation
  for NA-GA3C11, based on the estimated volume of the liquid crystal monomers,
  proposes that each layer consists of seven
  monomers\cite{resel_structural_2000}. Our best chance to answer this question
  is by using a molecular model orders of magnitude larger than any other
  reported atomistic liquid crystal membrane simulation, as we present here. We
  will directly change the layer composition and note its effect on membrane
  structure.

  \item How do monomers in each layer position themselves with respect to
  surrounding layers? \label{point:orientation}

  $\pi$-$\pi$ stacking interactions between aromatic monomer head groups may be
  a driving force of self-assembly in this system \cite{gazit_possible_2002}. Gas
  phase \textit{ab initio} studies of benzene dimers have shown a clear energetic
  advantage for parallel displaced and T-shaped $\pi$-$\pi$ stacking
  conformations over a sandwiched conformation \cite{sinnokrot_estimates_2002}.
  Substituted benzene rings exhibit an even stronger $\pi$-$\pi$ stacking
  attraction which favors the parallel displaced configuration in all cases
  except where the substitutions are extremely electron
  withdrawing\cite{waller_hybrid_2006,ringer_effect_2006}. In this study, we
  compare simulated X-ray diffraction patterns to experiment in order to deduce
  which stacking configuration is most likely. We also dive deeper into the
  origin of features present in the 2D diffraction patterns.

  \item Does our model support the existence of layers and if so, how well
  defined are the layers? \label{point:layers}

  %BJC: Reordered, so changing the wording here.
  Indications of strong $\pi$-$\pi$ stacking interactions in the direction
  perpendicular to the membrane plane, combined with the wedge-like character of
  the LLC monomers, suggests that monomers should arrange into disks that stack in
  layers. $\pi$-$\pi$ stacking will only occur between the aromatic
  monomer head groups which leaves no description of what is happening in the
  monomer tail region. The tails may entangle isotropically while maintaining
  stacking order among headgroups.

  \item Can the system exist in other metastable states or phases that are not
  accessed during experiments? \label{point:metastable}

  There remains the possibility that there is more than one metastable state
  associated with a given LLC phase. Simulating a membrane atomistically
  requires many atoms which limits the timescales acessible with MD. It is
  reasonable to expect that we will generate configurations which are kinetically
  trapped in a metastable free energy basin. We must be able to identify which
  state is produced experimentally.

  \item What constitutes a pore, and what are the details of its pore architecture? \label{point:poredefinition}

  The limited picture that experiment provides tells us that there are
  hexagonally packed, hydrophilic regions where transport is likely to occur.
  One may instinctively assume that these regions are hollow tube-like pathways.
  We will explore the composition of the pores and the partition between the
  hydrophilic and hydrophobic regions.

  \item Is it necessary to include any water in order to appropriately model
  the Col\textsubscript{h} phase? \label{point:water}

  While past work describes the Col\textsubscript{h} phase as dry~\cite{feng_scalable_2014}, 
  it has been suggested by experimentalists, in
  unpublished communications, that it is likely that neat monomer leaches small
  amounts of ambient water. Experimentally, achieving a hexagonal phase with a
  completely dry system has proven difficult. If neat monomer sits in ambient
  conditions, its color turns from transparent to slightly opaque and a hexagonal
  phase forms. Although we will not explore whether water is necessary for
  self-assembly, we hypothesize that the hydrogen bonding network formed by the
  water may play a role in structuring the pores and holding together the
  hexagonal phase. We can look at correlation functions and use simulated X-ray
  diffraction patterns to see if there is any meaningful structural difference
  between a ``dry'' and ``wet'' system.

@@

	\item \textit{R-$\pi$}: The location of the first is at $q_z$ = 1.7
	\AA$^{-1}$, corresponding to a real space separation of 3.7 \AA~. Previous
	work~\cite{feng_scalable_2014} attributes this reflection to $\pi$-$\pi$
	stacking between aromatic rings in the direction perpendicular to the membrane
	plane, or z-axis \cite{feng_scalable_2014}. For simplicity, we will refer to
	this reflection as R-$\pi$.

@@

	\item \textit{R-spots}: Within R-alkanes, are four spots of higher
	relative intensity.  Accordingly, we name these reflection R-spots. The
	location of all spots is $\sim 37$ degrees from the $q_z$ axis in their
	respective quadrants. In many liquid crystal systems one can explain the spots
	as the tilt angle of the alkane chains with respect to the membrane
	plane\cite{govind_simple_2001}.

@@

  The timescale for self-assembly of monomers into the hexagonal phase is
  unknown and likely outside of a reasonable length for an atomistic simulation,
  calling for a more efficient way to build the system. Previous work has shown
  a coarse-grained model self assemble into the H\textsubscript{II} phase
  configuration in $\sim$1000 ns \cite{mondal_self-assembly_2013}.  We
  attempted atomistic self-assembly by packing monomers into a box using Packmol
  \cite{martinez_packmol:_2009}. Simulations of greater than 100 ns show no
  indicators of progress towards an ordered system. To bypass the slow
  self-assembly process, we use Python scripts to assemble monomers into a
  structure close to one of a number of hypothesized equilibrium configurations
  (Figure~\ref{fig:python}).

  \begin{figure}
	\centering
	\includegraphics[width=0.75\linewidth]{build.PNG} %BJC: put an xyz axis with the unit cell
	\caption{(a) We parameterized a single monomer and annealed it to produce a low energy
		configuration. (b) A python script rotates and assembles monomers into layers with 
		hydrophilic centers. (c) We chose to stack twenty layers on top of each other to create
		pores. (d) The pores are duplicated and placed into a monoclinic unit cell.}\label{fig:python}
  \end{figure}

@@

  When constructing an initial configuration, there are a number of variables
  which require careful consideration while placing monomers. The equilibrium
  configuration is sensitive to some while insensitive to others. The starting
  pore radius, defined as the distance of a chosen head group carbon from the
  pore's central axis, does not influence the equilibrium structure if one choses
  a reasonable value. The pore radius is chosen to be 0.5 nm in our initial
  configurations. The initial distance between pores also has little effect on
  the the equilibrated structure However, one should not start them too close or
  there will be high energy repulsions during early equilibration. We chose an
  initial pore spacing of 4.5 nm, $\sim$10\% larger than the experimental value
  of 4.12 nm.  A sensitivity analysis of both parameters is presented in
  the supplemental information.  The distance between layers, the rotation of the
  layers with respect to adjacent layers, and the number of monomers per layer do
  influence the equilibrium structure and require further justification for their
  choices. We rely on experimental data to inform them.

@@

  We chose the relative interlayer orientation based on clues from diffraction
  data as well as the various known stacking modes of benzene and substituted
  benzene rings: sandwiched, parallel-displaced and T-shaped
  \cite{sinnokrot_estimates_2002}. We ruled out the T-shaped configuration
  because its $\sim$5 \AA~equilibrium stacking distance
  \cite{sinnokrot_estimates_2002} is inconsistent with R-$\pi$. It is also
  infeasible for the monomers to orient in the T-shaped conformation because of
  the bulky tail groups. We will explore the system's preference towards the
  sandwiched vs. parallel displaced stacking modes in some detail. Both have
  reported stacking distances near the R-$\pi$ value of 3.7 \AA. Headgroups in
  our sandwiched initial configuration stack directly on top of each other while
  headgroups in the parallel displaced initial configuration stack with an offset
  of $180/nmon$ degrees where $nmon$ equals the number of monomers per layer. See
  the Supplemental Information for a detailed illustration of the initial
  configurations in each mode.

@@

  In order to create a wet system, we solvated an initial configuration with
  water using \texttt{gmx solvate}. We remove all water molecules placed outside
  the pore region. Then we randomly remove water molecules inside the pore region
  until the pores reach the desired concentration of water. The remainder of the
  equilibration follows the same procedure as the dry system.

@@

  We calculated the nematic order parameter for our system in order to
  understand the degree of ordering among monomer head groups. Typically, the
  nematic order parameter is calculated for nematic liquid crystal systems which
  are characterized by unidirectional ordering of liquid crystal monomers. The
  preferred direction of monomers is defined by the unit director vector,
  $\mathbf{n}$. Assuming a single preferred direction of alignment, the nematic
  order parameter, $S$, is defined as:
  \begin{equation}
	 S = \frac{1}{2} \langle(3cos^2\theta -1)\rangle
	\label{eqn:nematic_order_parameter}
  \end{equation}
  where $\theta$ is the angle between the molecular long axis and $\mathbf{n}$.
  In a perfectly ordered system, the molecular axis should be aligned with
  $\mathbf{n}$ and give an order parameter of $S=1$. We are interested in
  quantifying the degree of monomer head group alignment between systems. We use
  Equation~\ref{eqn:nematic_order_parameter} to accomplish this by defining
  $\mathbf{n}$ as the z-axis (or pore axis), and then measuring the angles,
  $\theta$, between $\mathbf{n}$ and the vectors perpendicular to the plane of
  the aromatic head groups. See the supplemental information for a diagram.

@@

  Our simulations best support a model built with 5 monomers per layer based on
  the measured equilibrated pore-to-pore distances. To discern the composition of
  the monomer layers, addressing question (\ref{point:monomernum}), we ran
  simulations of systems created with 4 - 8 monomers per layer. We built systems
  in both the parallel displaced and sandwiched configurations with layers
  initially spaced 3.7 \AA~apart. We prepared equilibrated configurations
  according to the dry equilibration procedure. All systems are stable after 400
  ns of simulation. In a sense, all systems are at least metastable, addressing
  question (\ref{point:metastable}), however not all will make physical sense or fit
  the experimental profile that we are trying to match. Figure~\ref{fig:p2p}
  shows the pore spacing for all systems tested. Systems built with 5 monomers in
  each layer equilibrate to a pore spacing that is most consistent with the
  experimental value of 4.12 nm derived from SAXS measurements
  (Figure~\ref{fig:SAXS}). The remainder of this discussion will focus on the
  analysis of systems built with 5 monomers per layer.

  \begin{figure}
	\centering
	\includegraphics[width=\linewidth]{p2p.png}
	\caption{When we build systems with 5 monomers per layer , the
		equilibrated pore spacing is closest to the experimental value of 4.12 nm. The
			equilibrated pore spacing of the model increases as the number of monomers in
			each layer increases. The pore spacing of systems starting in the sandwiched
			configuration is systematically lower than those started in the parallel
			displaced configuration.}~\label{fig:p2p}
  \end{figure}

@@

  The parallel displaced configuration results in a simulated XRD pattern with
  the closest match to experiment. It produces the only pattern that exhibits all
  major reflections. R-alkanes and R-pores appear as they do in the sandwiched
  configuration. R-spots and R-$\pi$ appear, however with a lower intensity
  relative to R-alkanes when compared to the sandwiched configuration. R-helix
  appears due to the parallel displaced aromatic rings. It is a subharmonic of
  R-$\pi$ since the nearest vertically stacked head group to any given head group
  is $2\times$R-$\pi$~away.

>>>>

%MRS5: could be punchier. I will keep thinking.  I should be the last author.  Chen the names that Joe and Matt wanttu use (presumably with some middle initial?
\title{Understanding the nanoscale structure of hexagonal phase lyotropic 
liquid crystal membranes}
\author{Benjamin J. Coscia}
\author{Richard D. Noble}
\author{Douglas L. Gin}
\affiliation{Department of Chemical and Biological Engineering, University of Colorado Boulder, Boulder, CO 80309, USA}
\alsoaffiliation{Department of Chemistry and Biochemistry, University of Colorado Boulder, Boulder CO 80309, USA}
\author{Joe Yelk}
\author{Matthew Glaser}
\affiliation{Department of Physics, University of Colorado Boulder, Boulder CO, 80309, USA}
\author{Xunda Feng}
\affiliation{Department of Chemical and Environmental Engineering, Yale University, New Haven, Connecticut 06511, USA}
\author{Michael R. Shirts}
\email{michael.shirts@colorado.edu}
\affiliation{Department of Chemical and Biological Engineering, University of Colorado Boulder, Boulder, CO 80309, USA}
\begin{document}

@@

%MRS6: I would keep a paragraph about here how current membranes DON'T
%work for this purpose. But one paragraph is enough.x

@@

  Crosslinked lyotropic liquid crystal (LLC) membranes may be capable of
  performing highly selective separations. LLCs are amphiphilic molecules that
  have the ability to self-assemble into porous nanostructures
  \cite{smith_ordered_1997} which can be crosslinked to create mechanically
  strong membrane films with pores on the order of 1 nm in diameter
  \cite{zhou_supported_2005}. Unlike commercial NF membranes, LLC membrane pores
  are uniform in size. Since LLC membranes lack a pore size distribution, they
  inherently exhibit high selectivity due to their strict molecular weight
  cut-off (MWCO)~\cite{zhou_supported_2005}. Additionally, the LLC monomers
  %MRS6:
  described in this paper
  %chosen to make membranes for this application 
  are salts, and therefore lead to
  Donnan exclusion of ions in solution. The membrane gains a net surface charge
  when counterions from head groups that line the pore walls escape into the feed
  solution in order to balance the gradients of concentration and electric
  potential \cite{donnan_theory_1995}.

@@

  The H\textsubscript{II} phase pore geometry is 
%MRS6:
%better 
has a higher theoretical capacity
for transport than the
  Q\textsubscript{I} phase. The H\textsubscript{II} phase forms at room
  temperature in the presence of c.a.~10 wt\% water and consists of hexagonally
  packed, hydrophilic pore columns\cite{smith_ordered_1997}. In the absence of
  water, neat monomer will form the same hexagonal columnar structure which, in
  literature, has been referred to as the Col\textsubscript{h} thermotropic
  phase\cite{feng_scalable_2014} (See Figure~\ref{fig:hII_phase}).  The most
  promising Q\textsubscript{I} phase for membrane applications forms at 70\degree
  C when monomer is mixed with c.a.~20 wt\%
  glycerol\cite{carter_glycerol-based_2012}. Q\textsubscript{I} phase membranes
  consist of a tortuous network of three dimensionally interconnected pores that
  prevent optimal through-plane transport. The densely packed, non-tortuous and
  uniform sized pores of H\textsubscript{II} phase membranes represents the ideal
  geometry for achieving high solute flux\cite{matyka_tortuosity-porosity_2008}.
  Despite the promise of the H\textsubscript{II} phase, the hexagonally packed
  liquid crystalline domains, formed during self-assembly of NA-GA3C11 monomers, are
  isotropically aligned which is detrimental to membrane performance.

@@

  Our current understanding of 
%MRS6:
the molecular details of
LLC membrane 
%systems 
nanostructure
is not 
%MRS6: 
%rich enough to be
sufficient
  able to precisely design them for specific separations. Over the past 20 years,
  H\textsubscript{II}-phase LLC polymer membrane studies have been limited
  primarily to Na-GA3C11 with some characterization done after minor structural
  modifications. Resel et al.~varied the length of the monomer tails and the
  counterion used and observed its affect on pore spacing
  \cite{resel_structural_2000}. In a later study of rejection performance, it was
  shown that membranes formed by cross-linked Na-GA3C11 in the
  H\textsubscript{II} phase cannot separate solutes less than 1.2 nm in diameter
  because the pores are too large \cite{zhou_supported_2005}. We do not yet
  understand how to controllably reduce the effective pore size or how to tune
  the chemical environment in the nanopores of this or related materials for
  small molecule separations. The only source of predictive modeling for LLC
  systems have been macroscopic models that likely do not adequately describe
  transport at these length scales \cite{hatakeyama_water_2011}. It will be
  challenging to efficiently narrow down the large design space in a laboratory
  setting without a robust model.

@@

  \begin{figure}
  \centering
	\begin{subfigure}{0.45\linewidth}
		\centering
		\includegraphics[width=\textwidth]{cartoon_pore.png}
		\caption{}~\label{fig:undetailed_pore}
	\end{subfigure}
	\begin{subfigure}{0.45\linewidth}
		\centering
		\includegraphics[width=\textwidth]{detailed_pore.png}
		\caption{}~\label{fig:detailed_pore}
	\end{subfigure}
%MRS6: this is one place where I might go to passive voice, since anyone can only speculate not just us.  
  \caption{We can only speculate about solute behavior inside the membrane
	  pores based on our previous picture of the pore structure (a). 
%MRS6: should be present text, not ``will use'' since it's what's happening in this paper. Look for this throughout.
          We will use a
	  detailed molecular model in order to appropriately model the pore's complex
	  architecture which is crucial to understanding the mechanism of solute
	  transport (b). The head group region is colored blue
%MRS6: probably mention from actual simulations we will describe below.
	  and the tail region is colored cyan in both representations.}~\label{fig:detail}
  \end{figure}

@@

%MRS6: formatting suggestion
  \item \textit{If layers do exist, how many monomers constitute a single layer? \label{point:monomernum}} 
  A simple molecular simulation study of a similar molecule suggested that
  there are 4 monomers in each layer \cite{zhu_methacrylated_2006}. Their model
  likely does not sufficiently imitate the chemical environment present in the
  real system since it only contains 16 total monomers. A separate calculation
  for NA-GA3C11, based on the estimated volume of the liquid crystal monomers,
  proposes that each layer consists of seven
  monomers\cite{resel_structural_2000}. Our best chance to answer this question
  is by using a molecular model orders of magnitude larger than any other
  reported atomistic liquid crystal membrane simulation, as we present here. We
  will directly change the layer composition and note its effect on membrane
  structure.

  \item \textit{How do monomers in each layer position themselves with respect to
  surrounding layers? \label{point:orientation}}
  $\pi$-$\pi$ stacking interactions between aromatic monomer head groups may be
  a driving force of self-assembly in this system \cite{gazit_possible_2002}. Gas
  phase \textit{ab initio} studies of benzene dimers have shown a clear energetic
  advantage for parallel displaced and T-shaped $\pi$-$\pi$ stacking
  conformations over a sandwiched conformation \cite{sinnokrot_estimates_2002}.
  Substituted benzene rings exhibit an even stronger $\pi$-$\pi$ stacking
  attraction which favors the parallel displaced configuration in all cases
  except where the substitutions are extremely electron
  withdrawing\cite{waller_hybrid_2006,ringer_effect_2006}. In this study, we
  compare simulated X-ray diffraction patterns to experiment in order to deduce
  which stacking configuration is most likely. We also dive deeper into the
  origin of features present in the 2D diffraction patterns.

  \item \textit{Does our model support the existence of layers and if so, how well
  defined are the layers? \label{point:layers}} 
  %BJC: Reordered, so changing the wording here.
  Indications of strong $\pi$-$\pi$ stacking interactions in the direction
  perpendicular to the membrane plane, combined with the wedge-like character of
  the LLC monomers, suggests that monomers should arrange into disks that stack in
  layers. $\pi$-$\pi$ stacking will only occur between the aromatic
  monomer head groups which leaves no description of what is happening in the
  monomer tail region. The tails may entangle isotropically while maintaining
  stacking order among headgroups.

  \item \textit{Can the system exist in other metastable states or phases that are not
  accessed during experiments? \label{point:metastable}}
  There remains the possibility that there is more than one metastable state
  associated with a given LLC phase. Simulating a membrane atomistically
  requires many atoms which limits the timescales acessible with MD. It is
  reasonable to expect that we will generate configurations which are kinetically
  trapped in a metastable free energy basin. We must be able to identify which
  states 
  %MRS6: best not to set up too high expectations
  %is produced experimentally.
  are most consistent with experiment.

  \item \textit{What constitutes a pore, and what are the details of its pore architecture? \label{point:poredefinition}}
  The limited picture that experiment provides tells us that there are
  hexagonally packed, hydrophilic regions where transport is likely to occur.
  One may instinctively assume that these regions are hollow tube-like pathways.
  We will explore the composition of the pores and the partition between the
  hydrophilic and hydrophobic regions.

  \item \textit{Is it necessary to include any water in order to appropriately model
  the Col\textsubscript{h} phase? \label{point:water}}
  While past work describes the Col\textsubscript{h} phase as dry~\cite{feng_scalable_2014}, 
  it has been suggested by experimentalists, in
  unpublished communications, that it is likely that neat monomer leaches small
  amounts of ambient water. Experimentally, achieving a hexagonal phase with a
  completely dry system has proven difficult. If neat monomer sits in ambient
  conditions, its color turns from transparent to slightly opaque and a hexagonal
  phase forms. Although we will not explore whether water is necessary for
  self-assembly, we hypothesize that the hydrogen bonding network formed by the
  water may play a role in structuring the pores and holding together the
  hexagonal phase. We can look at correlation functions and use simulated X-ray
  diffraction patterns to see if there is any meaningful structural difference
  between a ``dry'' and ``wet'' system.

@@

	\item \textit{R-$\pi$}: The location of the first is at $q_z$ = 1.7
	\AA$^{-1}$, corresponding to a real space separation of 3.7 {\AA}. Previous
	work~\cite{feng_scalable_2014} attributes this reflection to $\pi$-$\pi$
	stacking between aromatic rings in the direction perpendicular to the membrane
	plane, or z-axis \cite{feng_scalable_2014}. For simplicity, we will refer to
	this reflection as R-$\pi$.

@@

	\item \textit{R-spots}: Within R-alkanes, are four spots of higher
	relative intensity.  Accordingly, we name these reflection R-spots. The
	location of all spots is $\sim\;37^{\circ}$ from the $q_z$ axis in their
	respective quadrants. In many liquid crystal systems one can explain the spots
	as the tilt angle of the alkane chains with respect to the membrane
	plane\cite{govind_simple_2001}.

@@

  The timescale for self-assembly of monomers into the hexagonal phase is
  unknown and likely outside of a reasonable length for an atomistic simulation,
  calling for a more efficient way to build the system. Previous work has shown
  a coarse-grained model self assemble into the H\textsubscript{II} phase
  configuration in $\sim$1000 ns \cite{mondal_self-assembly_2013}.  We
  attempted atomistic self-assembly by packing monomers into a box using Packmol
  \cite{martinez_packmol:_2009}. Simulations of greater than 100 ns show no
  indicators of progress towards an ordered system. To bypass the slow
  self-assembly process, we use Python scripts to assemble monomers into a
  structure close to one of a number of hypothesized equilibrium configurations
  (Figure~\ref{fig:python}).
  %MRS6: question; we would like to make these releasable in GitHub.
  \begin{figure}
	\centering
	\includegraphics[width=0.75\linewidth]{build.PNG} %BJC: put an xyz axis with the unit cell
	\caption{(a) We parameterized a single monomer and annealed it to produce a low energy
		configuration. (b) A Python script rotates and assembles monomers into layers with 
		hydrophilic centers. (c) We chose to stack twenty layers on top of each other to create
		pores. (d) The pores are duplicated and placed into a monoclinic unit cell.}\label{fig:python}
  \end{figure}

@@

  When constructing an initial configuration, there are a number of variables
  which require careful consideration while placing monomers. The equilibrium
  configuration is sensitive to some while insensitive to others. The starting
  pore radius, defined as the distance of a chosen head group carbon from the
  pore's central axis, does not influence the equilibrium structure if one choses
  a reasonable value. The pore radius is chosen to be 0.5 nm in our initial
  configurations. 
  %MRS6: added.
  The initial distance between pores, within a wide range, also has little effect on
  the the equilibrated structure. However, one should not start them too close or
  there will be high energy repulsions during early equilibration. We chose an
  initial pore spacing of 4.5 nm, $\sim$10\% larger than the experimental value
  of 4.12 nm.  A sensitivity analysis of both parameters is presented in
  the 
  %MRS6: check for J phys chem B to see if its supporting or supplemental information or material.
  supplemental information.  The distance between layers, the rotation of the
  layers with respect to adjacent layers, and the number of monomers per layer do
  influence the equilibrium structure and require further justification for their
  choices. We rely on experimental data to inform them.

@@

  We chose the relative interlayer orientation based on clues from diffraction
  data as well as the various known stacking modes of benzene and substituted
  benzene rings: sandwiched, parallel-displaced and T-shaped
  \cite{sinnokrot_estimates_2002}. We ruled out the T-shaped configuration
  because its $\sim$5 \AA~equilibrium stacking distance
  \cite{sinnokrot_estimates_2002} is inconsistent with R-$\pi$. It is also
  infeasible for the monomers to orient in the T-shaped conformation because of
  the bulky tail groups. We will explore the system's preference towards the
  sandwiched vs. parallel displaced stacking modes in some detail. Both have
  reported stacking distances near the R-$\pi$ value of 3.7 \AA. Headgroups in
  our sandwiched initial configuration stack directly on top of each other while
  headgroups in the parallel displaced initial configuration stack with an offset
  of $180^\circ/nmon$ where $nmon$ is the number of monomers per layer. See
  the Supplemental Information for a detailed illustration of the initial
  configurations in each mode.

@@

  %MRS6: some parameters missing here (time run between insertions, etc). 
  %I assume they are in supporting information?
  In order to create a ``wet'' system, we solvated an initial configuration with
  water using \texttt{gmx solvate}. We remove all water molecules placed outside
  the pore region. Then we randomly remove water molecules inside the pore region
  until the pores reach the desired concentration of water. The remainder of the
  equilibration follows the same procedure as the dry system.

@@

  We calculated the nematic order parameter for our system in order to
  understand the degree of ordering among monomer head groups. Typically, the
  nematic order parameter is calculated for nematic liquid crystal systems which
  are characterized by unidirectional ordering of liquid crystal monomers. The
  preferred direction of monomers is defined by the unit director vector,
  $\mathbf{n}$. Assuming a single preferred direction of alignment, the nematic
  order parameter, $S$, is defined as:
  \begin{equation}
	 S = \frac{1}{2} \langle(3\cos^2\theta -1)\rangle
	\label{eqn:nematic_order_parameter}
  \end{equation}
  where $\theta$ is the angle between the molecular long axis and $\mathbf{n}$.
  In a perfectly ordered system, the molecular axis should be aligned with
  $\mathbf{n}$ and give an order parameter of $S=1$. We are interested in
  quantifying the degree of monomer head group alignment between systems. We use
  Eq.~\ref{eqn:nematic_order_parameter} to accomplish this by defining
  $\mathbf{n}$ as the z-axis (or pore axis), and then measuring the angles,
  $\theta$, between $\mathbf{n}$ and the vectors perpendicular to the plane of
  the aromatic head groups. See the supplemental information for a diagram.

@@

  Our simulations best support a model built with 5 monomers per layer based on
  the measured equilibrated pore-to-pore distances. To discern the composition of
  the monomer layers, addressing question (\ref{point:monomernum}), we ran
  simulations of systems created with 4--8 monomers per layer. We built systems
  in both the parallel displaced and sandwiched configurations with layers
  initially spaced 3.7 \AA~apart. We prepared equilibrated configurations
  according to the dry equilibration procedure. All systems are stable after 400
  ns of simulation. 
  %MRS6: Above, probably need to be a bit more clear what you mean by stable (no noticable drift in pore-to-pore), and if you are only using data after the point of stability.
In a sense, all systems are at least metastable, addressing
  question (\ref{point:metastable}), however not all will make physical sense or fit
  the experimental profile that we are trying to match. Figure~\ref{fig:p2p}
  shows 
%MRS6:
%the pore spacing 
the equilibrated pore-to-pore distances 
for all systems tested. 
Systems built with 5 monomers in
  each layer equilibrate to a pore spacing that is most consistent with the
  experimental value of 4.12 nm derived from SAXS measurements
  (Figure~\ref{fig:SAXS}). 
%MRS6: discuss very briefly how far off 4 and 6 are to motivate selecting 5 exclusively.
The remainder of this discussion will focus on the
  analysis of systems built with 5 monomers per layer.

  \begin{figure}
	\centering
	\includegraphics[width=\linewidth]{p2p.png}
        %MRS6: should there be a grapical demonstration of sandwiched vs. parallel displaced, especially equilbrated?
        %MRS6: some tweaks below.
	\caption{Systems with 5 monomers per layer have
		equilibrated pore spacing closest to the experimental value of 4.12 nm. The
			equilibrated pore spacing of the model increases as the number of monomers in
			each layer increases. The pore spacing of systems starting in the sandwiched
			configuration is systematically lower than those started in the parallel
			displaced configuration.}~\label{fig:p2p}
  \end{figure}

@@

  The parallel displaced configuration results in a simulated XRD pattern with
  the closest match to experiment. It produces the only pattern that exhibits all
  major reflections. R-alkanes and R-pores appear as they do in the sandwiched
  configuration. R-spots and R-$\pi$ appear, however with a lower intensity
  relative to R-alkanes when compared to the sandwiched configuration. R-helix
  appears due to the parallel displaced aromatic rings. It is a subharmonic of
  R-$\pi$ since the nearest vertically stacked head group to any given head group
  is $2\times$R-$\pi$~away. 
  %BJC: adjust figure size and alignment -- probably easiest to set figure size in matplotlib
  \begin{figure}
  \begin{subfigure}{0.3\linewidth}
        \centering
        \vspace{-0.2em}
        \includegraphics[width=1.1\linewidth,trim={1cm 0 1.3cm 0},clip]{offset_rzplot.png}
        \caption{}~\label{fig:rz_offset}
  \end{subfigure}
  \begin{subfigure}{0.3\linewidth}
        \centering
%        \vspace{0.25em}
        \includegraphics[scale=0.285]{WAXS_raw.png}
        \caption{}~\label{fig:raw_waxs}
  \end{subfigure}
  \begin{subfigure}{0.3\linewidth}
        \centering
        \includegraphics[width=1.1\linewidth,trim={1cm 0 1.3cm 0},clip]{layered_rzplot.png}
        \caption{}~\label{fig:rz_layered}
  \end{subfigure}
  \begin{subfigure}{0.0544\linewidth}
        \centering
        \vspace{-4.00em}
        \includegraphics[width=\linewidth]{colorbar_seismic.png}
  \end{subfigure}
%MRS6: question: the SAXS region looks different between sandwiched and parallel displaced. Any thoughts why? We can discuss in person.  And seems like have to comment on the difference in intensity between Sandwich and parallel displaced in the pi-pi region.
%MRS6: I wonder to what extent showing the reflections of tails removed would make it clearer pi-pi is really there.  Maybe in supporting info?
  \caption{The simulated XRD pattern generated from the equilibrated trajectory
	  of the parallel displaced configuration (a) exhibits all major reflections
	  present in the experimental WAXS pattern (b). The simulated XRD pattern
	  generated from the equilibrated trajectory in the sandwiched configuration (c) shows
	  all major reflections except R-helix.}
  \label{fig:XRDsim}
  \end{figure}
